---
title: "College Survey 2014 summary"
author: "weiran"
date: "June 21, 2016"
output: html_document
---


```{r,message=FALSE,echo=FALSE,warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(grid)
library(gridExtra)
library(lubridate)
library(tigerstats)

setwd("C:\\Users\\weiran\\Documents\\HHSurvey\\college")

```

#Data cleannig
this section read data and change the type of data

```{r,warning=FALSE}

# read data
df.person<- read.csv("1_PSRC_College_PersonDataset_2015-1-23_V1-3.csv")
df.trip<- read.csv("2_PSRC_College_TripDataset_2015-1-23_v2-1_LINKED.csv")


cols <- c("college")
df.trip[,cols] <- data.frame(apply(df.trip[cols], 2, as.factor))
levels(df.trip$college) <- c("Bellevue.C","Everett.C", "Green River.C", "Seattle.C", "U.W")

cols1 <- c("college")
df.person[,cols1] <- data.frame(apply(df.person[cols1], 2, as.factor))
levels(df.person$college) <- c("Bellevue.C","Everett.C", "Green River.C", "Seattle.C", "U.W")

df.live <- df.person %>%
  select(personID=personid,live,h_district_name,h_districtnum,home_cnty)

df <- df.trip %>%
  mutate(mode=as.factor(mode),
         vehicle=as.factor(vehicle)) %>%
  mutate(traveldate = dmy(as.character(traveldate))) %>%
  mutate(d_purpose = as.factor(d_purpose),
         o_purpose = as.factor(o_purpose)) %>%
  mutate(trip_commute = ifelse((o_purpose==1 & d_purpose==2)|
                                 (o_purpose==2 & d_purpose ==1),
                               1,0)) %>%
  left_join(df.live,by='personID') %>%    #join person dataset variable
  mutate(add_commute=ifelse(live==2,'off campus','on campus')) %>%
  mutate(trip_commute=as.factor(trip_commute),
         add_commute=as.factor(add_commute)) %>%
  mutate(home_cnty = as.factor(home_cnty)) 



levels(df$home_cnty) <- c('King','Kitsap','Pierce','Snohomish','Other WA cnty','Outside WA')

```

## Data describe
this describe the sample size of the data among college and campus\\

campus code
1	Seattle campus
2	Bothell campus
3	Tacoma campus
4	North Seattle College
5	Seattle Central College
6	South Seattle College (Main Campus)
7	Main campus
8	Enumclaw campus
9	Kent campus
10	Downtown Auburn Center
11	Main campus
12	North campus
13	Online only - I never go to campus
14	South Seattle College (Georgetown Campus)
15	Seattle Vocational Institute

```{r,warning=FALSE}
table(df.trip$college)
table(df.person$college)
table(df.person$campus)
```
## 1 Mode choice
mode code:
1	Drove alone
3	Drove/rode with people not in household
4	Motorcycle/moped/scooter
5	Vanpool
6	Bicycle
7	Walk, jog, or wheelchair
8	Bus (public transit)
9	Train (rail and monorail)
10	Ferry or water taxi
11	Streetcar
13	Taxi or other hired car service (e.g. Lyft, Uber)
14	Paratransit
16	Airplane or helicopter
15	Private bus or shuttle
17	Other (e.g. skateboard, kayak, motor home, etc.)


```{r,warning=FALSE}
rowPerc(xtabs(~df$mode))
rowPerc(xtabs(~df[df$trip_commute==1,]$mode)) #commute trip modeshare


ggplot(df, aes(x=factor(mode)))+ geom_bar()+
  facet_grid(college~.,scale="free")

#commute trip modeshare
ggplot(df[df$trip_commute==1,], aes(x=factor(mode)))+ geom_bar()+
  facet_grid(college~.,scale="free")

```
## Trips per day, total  and by purpose


```{r,warning=FALSE}
# trips stat summary
df.tripstat<- df.trip %>%
  mutate(traveldate=dmy(as.character(traveldate))) %>%
  distinct()%>%
  group_by(college,traveldate) %>%
  summarise(totaltrip.coldate = mean(total_trip))

# box plot for daily trips made by college. Senseless to summarize total trips, 
#since for each school different number of students are sampled
ggplot(df.tripstat,aes(college,totaltrip.coldate)) + geom_boxplot() + ylab('average daily trips made by each student')
# # box plot for each person each day
ggplot(df.trip,aes(x=college,y=total_trip))+geom_boxplot()+
  geom_hline(aes(yintercept=mean(total_trip, na.rm=T)),  
             color="red", linetype="dashed", size=1)

df.trippurpose <- df%>%
  group_by(college,d_purpose) %>%
  summarise(numtrips=n())
  
ggplot(df.trippurpose,aes(d_purpose,numtrips)) + geom_bar(stat='identity') +
  facet_grid(college~.,scale="free") + xlab('trip purpose') + ylab('num. of trips') 


```

##3. trip length summary
purpose code

1	Go home/to dorm
2	Go to school (e.g. go to class, go to tutoring)
3	Go to work/job
4	Go to other work-related place (e.g. meeting, second job, delivery)
5	Go grocery shopping
6	Go to other shopping (e.g. mall, pet store)
7	Go to medical appointment (e.g. doctor, dentist)
8	Conduct personal business (e.g. bank, post office)
9	Drop off/pick up someone (e.g. drop off friend at bus stop)
10	Go exercise (e.g. gym, walk, jog, bike ride)
11	Go to restaurant to eat/get take-out
12	Attend social event (e.g. visit with friends, family, co-workers)
13	Attend recreational event (e.g. movies, sporting event)
14	Go to religious/community/volunteer activity
15	Transfer to another mode of transportation (e.g. change from ferry to bus)
16	Other
17	Go to school-related activity

```{r,warning=FALSE}

# Trip length/distance frequency and overall
ggplot(df,aes(trip_dur_reported)) +geom_histogram(colour="black", fill="white") +
  geom_vline(aes(xintercept=mean(trip_dur_reported, na.rm=T)),  
             color="red", linetype="dashed", size=1)+
  # facet_grid(college~.,scales='free_y') + 
  xlim(0,150) + xlab('trip length (in minutes)') + ylab('frequency')


# trip length by mode

# trip length by purpose
ggplot(df,aes(as.factor(d_purpose),trip_dur_reported)) + geom_boxplot() +
  # facet_grid(college~.,scales='free_y') + 
  # scale_y_log10()+
  ylim(0,75) + 
  ylab('trip length (in minutes)') + xlab('trip purpose')

# commute trip length by college
ggplot(df[df$trip_commute==1,],aes(college,trip_dur_reported)) + geom_boxplot() +
  geom_hline(aes(yintercept=mean(trip_dur_reported, na.rm=T)),  
             color="red", linetype="dashed", size=1)+
  ylim(0,75) + 
  ylab('trip length (in minutes)') + xlab('college')

ggplot(df[df$trip_commute==1,],aes(college,gdist)) + geom_boxplot() +
  geom_hline(aes(yintercept=mean(gdist, na.rm=T)),  
             color="red", linetype="dashed", size=1)+
  ylim(0,30) + 
  ylab('trip distance (miles)') + xlab('college')

# distinct start and destination
# df.destinfo can be used to define commuters/non-commuters.
df.destinfo <- df.trip %>%
  select(place_start,place_end) %>%
  group_by(place_start,place_end) %>%
  summarise(cnt = n()) %>%
  # group_by(place_start) %>%
  ungroup() %>%
  arrange(desc(cnt)) %>%
  filter(place_end == 'HOME (OFF-CAMPUS)')  
```

```{r,warning=FALSE}

```

```{r,warning=FALSE}

```


## 4. VMT per day per person
```{r,warning=FALSE}
#clean
df<- df%>%
  filter(mode!=16)#exclude flight trips
levels(df$h_district_name)[levels(df$h_district_name)==""] <- NA

# by college
df4.1 <- df %>%
  filter(driver==1,mode!=10) %>%
  group_by(personID,college) %>%
  summarise(vmtperday = sum(gdist)/n_distinct(traveldate))

ggplot(df4.1,aes(college,vmtperday)) + geom_boxplot() +
  geom_hline(aes(yintercept=mean(vmtperday, na.rm=T)),  
             color="red", linetype="dashed", size=1)+
  scale_y_continuous(breaks = seq(0,50,5),limits = c(0,50))+
  ylab('VMT per person per day') + xlab('college')

# by commuter

df4.2 <- df %>%
  filter(driver==1,trip_commute==1) %>%
  group_by(personID,college) %>%
  summarise(vmtperday = sum(gdist)/n_distinct(traveldate))

ggplot(df4.2,aes(college,vmtperday)) + geom_boxplot() +
  geom_hline(aes(yintercept=mean(vmtperday, na.rm=T)),  
             color="red", linetype="dashed", size=1)+
  scale_y_continuous(breaks = seq(0,40,5))+
  ylab('VMT of commuting per person per day ') + xlab('college')


```

#5. Time of day travel distribution for different schools vs 

```{r,warning=FALSE}
# regiondal distributions (hist of departure time by time of day)
df5 <- df %>%
  select(time_start_hhmm,college,cnty_start,cnty_end,trip_commute) %>%
  mutate(time_start = hm(as.character(time_start_hhmm))) %>%
  mutate(hour=hour(time_start),minute=minute(time_start)) %>%
  mutate(time=hour+minute/60) %>%
  mutate(cnty_start=as.factor(cnty_start),
         cnty_end=as.factor(cnty_end))
class(df5$time) <- 'POSIXct'

# by school
ggplot(df5,aes(time)) + geom_histogram(aes(fill=college)) +
  scale_x_datetime(labels=date_format("%S:00")) +
  xlab('departure time') + ylab('histogram') + 
  facet_grid(college~.,scales='free_y')

# by county start?
levels(df5$cnty_start) <- c('King','Kitsap',
                            'Pierce','Snohomish',
                            'Other WA cnty',
                            'Outside WA')
ggplot(df5,aes(time)) + geom_histogram(aes(fill=cnty_start)) +
  scale_x_datetime(labels=date_format("%S:00")) +
  xlab('departure time') + ylab('histogram') + 
  facet_grid(cnty_start~.,scales='free_y')

```

## 6. Commute times by home location 


```{r,warning=FALSE}
# (h_district_name)
df6 <- df %>%
  filter(trip_commute==1) %>%
  select(trip_dur_reported,h_district_name,college)

ggplot(df6,aes(h_district_name,
               trip_dur_reported)) +
  # geom_point()+
  # stat_summary(fun.data = "mean_cl_boot", colour = "red", size = .8)+
  geom_dotplot(binaxis="y",binwidth = .5,stackdir = "center") + 
  stat_summary(fun.data = "mean_cl_boot", colour = "red", size = .8)+
  xlab('home district') +ylim(0,150) +
  ylab('trip time (in min)')+ 
  facet_grid(college~.,scales='free_y')

```

## 7. carshare data vs. 2014 regional results

```{r,warning=FALSE}
df.person_c<- df.person%>%
  filter(!is.na(carshare_car2go))

car2go<- df.person_c%>%
  group_by(carshare_car2go)%>%
  summarise(Freq = n(),Freqw = sum(expwt))%>%
  mutate(Percent =percent(round(Freq/sum(Freq),digit=3)),Percent_w = percent(round(Freqw/sum(Freqw),digit=3)),
         member = carshare_car2go)

zipcar<-  df.person_c%>%
  group_by(carshare_zipcar)%>%
  summarise(Freq = n(),Freqw = sum(expwt))%>%
  mutate(Percent =percent(round(Freq/sum(Freq),digit=3)),Percent_w = percent(round(Freqw/sum(Freqw),digit=3)),
         member = carshare_zipcar)

relayrides<- df.person_c%>%
  group_by(carshare_relayrides)%>%
  summarise(Freq = n(),Freqw = sum(expwt))%>%
  mutate(Percent =percent(round(Freq/sum(Freq),digit=3)),Percent_w = percent(round(Freqw/sum(Freqw),digit=3)),
         member = carshare_relayrides)

other <- df.person_c%>%
  group_by(carshare_other)%>%
  summarise(Freq = n(),Freqw = sum(expwt))%>%
  mutate(Percent =percent(round(Freq/sum(Freq),digit=3)),Percent_w = percent(round(Freqw/sum(Freqw),digit=3)),
         member = carshare_other)
dontknow<- df.person_c%>%
  group_by(carshare_dontknow)%>%
  summarise(Freq = n(),Freqw = sum(expwt))%>%
  mutate(Percent =percent(round(Freq/sum(Freq),digit=3)),Percent_w = percent(round(Freqw/sum(Freqw),digit=3)),
         member = carshare_dontknow)

none<-df.person_c%>%
  group_by(carshare_none)%>%
  summarise(Freq = n(),Freqw = sum(expwt))%>%
  mutate(Percent =percent(round(Freq/sum(Freq),digit=3)),Percent_w = percent(round(Freqw/sum(Freqw),digit=3)),
         member = carshare_none)

type=c(rep("car2go",2),rep("zipcar",2),rep("relayrides",2),rep("other",2),rep("dontknow",2),rep("carshare",2))
carshare<-rbind(car2go[,-c(1,2,3)],relayrides[,-c(1,2,3)],zipcar[,-c(1,2,3)],none[,-c(1,2,3)],dontknow[,-c(1,2,3)],other[,-c(1,2,3)])
carshare<- cbind(carshare,type)
carshare

ggplot(df.person,aes(college,fill=factor(carshare_none))) +  geom_bar(position='dodge') +  xlab('home district') + ylab('carshare usage') +
  scale_fill_discrete(name='car share',labels=c('Member','Not member'))

```

## 8. Transit pass ownership and usage vs. region


```{r,warning=FALSE}
# transit percentage
rowPerc(xtabs(~df.person$transitpay_orca))
rowPerc(xtabs(~df.person$transitpay_cash))
rowPerc(xtabs(~df.person$transitpay_upass))
# 
# df_8 <- df.person %>%
#   mutate(transitpay_upass=as.factor(ifelse(is.na(transitpay_upass),0,1))) %>%
# 
ggplot(df.person,aes(college,fill=factor(transitpay_upass))) + geom_bar(position='dodge') + 
  scale_fill_discrete(name='transit pass',labels=c('No','Yes'))

```

## 9. Parking pass ownership and lot usage by district or school


```{r,warning=FALSE}
df9 <- df.person %>%
  filter(!is.na(parking_pass)) %>%
  mutate(parking_pass=as.factor(ifelse(parking_pass==1,'Yes','No'))) %>%
  mutate(live=as.factor(ifelse(live==1,'on campus','off campus')))

ggplot(df9[!is.na(df9$parking_pass),],aes(college,fill=parking_pass)) + geom_bar(position='dodge') 

```

## 10. walk/bike/transit frequency


```{r,warning=FALSE}
df10 <- df.person %>%
  mutate(transit_freq=as.factor(transit_freq),bike_freq=as.factor(bike_freq),walk_freq=as.factor(walk_freq)) 

rowPerc(xtabs(~df.person$walk_freq))
rowPerc(xtabs(~df.person$transit_freq))
rowPerc(xtabs(~df.person$bike_freq))

# walk freq.

ggplot(df.person,aes(as.factor(walk_freq))) + geom_bar(aes(fill=as.factor(walk_freq))) + 
  facet_grid(college~.,scales='free_y') + 
  xlab('Walk frequency (1-7:from high to low)') + ylab('counts') + 
  ggtitle('Walk freq. by college')

# transit freq.
ggplot(df10,aes(transit_freq)) + geom_bar(aes(fill=transit_freq)) + 
  facet_grid(college~.,scales='free_y') + 
  xlab('transit frequency (1-7:from high to low)') + ylab('counts') + 
  ggtitle('transit freq. by college')

# bike freq.
ggplot(df10,aes(bike_freq)) + geom_bar(aes(fill=bike_freq)) +
  facet_grid(college~.,scales='free_y') + 
  xlab('bike frequency (1-7:from high to low)') + ylab('counts') + 
  ggtitle('bike freq. by college')

```
## 11. driver's license vs region over all and by age groups (what is driver's license ownership of 


```{r,warning=FALSE}
# 18-24 old at UW main campus vs 18-24 at a community college, vs 18-24 year in the region over all)
df11 <- df.person %>%
  filter(!is.na(license)) %>%
  mutate(license=as.factor(ifelse(license==1,'Yes','No'))) %>%
  mutate(age_group = ifelse(age==1,'16-17',
                            ifelse(age %in% c(2,3,4), '18-24','>25')))

ggplot(df11,aes(college,fill=license)) + geom_bar(position = 'dodge') + facet_grid(age_group~.,scales = 'free_y')
ggplot(df11,aes(age_group,fill=license)) + geom_bar(position = 'dodge') 
  
```

## 12. taxi_type vs. region (regular cab vs. uber/lyft for college students vs. region)


```{r,warning=FALSE}
df12 <- df %>%
  filter(!is.na(taxi_type)) %>%
  mutate(taxi_type = as.factor(ifelse(taxi_type==1,'regular','uber/lyft'))) 

ggplot(df12,aes(college,fill=taxi_type)) + geom_bar(position='dodge') + ylab('taxi count')

```



```{r,warning=FALSE}

```



```{r,warning=FALSE}

```
